# Making raster data

## Generating rasters

Rasters are gridded data that are meant to represent a continuous variable over space. From the lecture, we saw that we can create a random raster by generating a matrix of random values of a given size. Here, we'll generate a 100x100 raster with values drawn from a standard normal distribution. 

```{r }
#Generate 10000 random values in a 10x10 matrix
randomValues<-matrix(rnorm(10000,0,1),nrow=100,ncol=100)

#Turn matrix into SpatRaster object
raster<-rast(randomValues)

#Plot raster
plot(raster)
```
In fact, we don't need to supply random values at all. This could be a completely empty raster if we omit the `rnorm` argument altogether.

```{r }
#Generate 10000 random values in a 10x10 matrix
noValues<-matrix(nrow=100,ncol=100)

#Turn matrix into SpatRaster object
emptyRaster<-rast(noValues)

#Plot raster
plot(emptyRaster)
```

Values in a raster, empty or otherwise, can be added or modified in a number of ways. One way we can plug in values to our raster is with the `values` function. In this case, we'll sort some random values before they get added by wrapping the `rnorm` call in the`sort` function. 

```{r}
#Add values to the empty raster
values(emptyRaster)<-sort(runif(10000,0,1))
#Plot the now not-so-empty raster
plot(emptyRaster)
```

## Non-numeric rasters

Since rasters typically represent continuous data, it makes sense that they often contain numeric values. However, this needn't always be the case. Here we'll make a raster of different landcover types. First, let's create a vector with some landcover types:

```{r}
#Land cover types
lctypes<-c("grassland","forest","shrubland")
```

Next, we'll generate a sample of 100 landcovers with the `sample` function:

```{r}
#Sample landcover types
landcover<-sample(lctypes,100,replace=TRUE)
#Look at the sample
landcover
```

Now we can create a 10x10 raster of landcover types using `rast`:

```{r}
#Make landcover raster
lcRaster<-rast(ncols=10,nrows=10,resolution = 1, xmin = 0, xmax = 10, ymin = 0, ymax = 10,vals=landcover)
#Plot landcover raster
plot(lcRaster)
```
Nice! Finally, we can also make rasters that have boolean, or TRUE/FALSE, values. Here, we'll make a raster that shows whether a cell in our landcover raster is forested or not. 

```{r}
#Make a true/false raster of forest cells
forestMask<-lcRaster=="forest"
#Take a look
forestMask
```

The values of this new raster are now only true or false. We can take a look at this in a plot.

```{r}
#Plot the forest true/false
plot(forestMask)
```

Boolean rasters can be especially useful as *masks*, or a layer that is used to include or omit certain cells from an operation. We'll look at how these work later on. 

## Try it yourself!

Try and do the following

* Generate a 20x20 matrix of 400 random uniform values between 0 and 100
* Turn this into a raster and plot it
* Use this to create a boolean raster of values greater than 80. 
* Create a multilayer raster from the two rasters you've made
